Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-03 manuscript/manuscript.tex
Finish introduction.

One of the most important and widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
The Kingman coalescent is a stochastic process that generates gene genealogies: trees representing the patterns of shared ancestry of sampled individuals.
Inference methods use these genealogies as latent variables linking demographic parameters to observable data~\autocite{RosenbergNordborg2002}.
The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{ms})---tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).

In its simplest form, the Kingman coalescent has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
Under a variety of conditions, the coalescent rate is inversely proportional to the size of the population~\autocite{Kingman1982b}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
This observation is the basis for a variety of methods to infer the population size history from genetic data.

With modern population genomics data sets, full-data likelihood models are impractical, so population-size inference is typically done on informative summary statistics.
One widely used statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in the sample.
The expected number of mutations at a given frequency depends on the branch lengths, and hence, on the coalescent rate.
Thus, one can infer the coalescent rate by integrating over tree topologies, weighted by their probabilities under the Kingman coalescent.
Some methods perform this integration by Monte Carlo simulation (e.g., \cite{CoventryEtAl2010, ExcoffierEtAl2013}).
Others (e.g., \cite{Nielsen2000}, \cite{BhaskarEtAl2015}), compute the expected site frequency spectrum directly for simple demographic models using the results of \cite{GriffithsTavare1998} or \cite{PolanskiKimmel2003}.
(Another class of SFS-based methods are based on forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)

[Question: where do I cite the SMC models?]

Model misspecification. Schrider, others on problems of inference. Evidence for widespread selection.

Class of non-Kingman coalescents: multiple mergers. Where they arise. What they mean for inference (e.g. distort SFS, long-range linkage, timescale issues)

Tests for multiple mergers and non-neutrality in general.
General neutrality tests (Tradition of Taj's D. Achaz)
Problem with confounding (compare to neutral background)
Local tests for sweeps (haplotype tests)
Specific tests for multiple mergers

Our test, based on the 2-site frequency spectrum:
\begin{itemize}
\item Genome-wide. Not comparing loci to putatively neutral background.
\item Distinguishes time-varying N from multiple mergers.
\item SNP data.
\item Doesn't require phasing or ancestral states.
\item Easy to compute.
\item Works in principle with small samples.
\end{itemize}

Outline:
\begin{itemize}
\item Define statistic, show that it works without recombination. (Detects multiple mergers, doesn't respond to population size changes)
\item Coalescent simulations with recombination (it still works).
\item Forward-time simulations of linked sweeps (still works).
\item Application to D. mel (it detects non-Kingman)
\end{itemize}

Modern population genetics makes it possible to learn the evolutionary and demographic history of a population from its contemporary genetic diversity~\autocite{Hahn2018}.
The genetic differences among individuals reflect their genealogical history, which, in turn, reflects the history of various population parameters such as the population size, spatial structure, and the influence of natural selection.
By modeling how the (unobserved) genealogical process depends on these parameters and how patterns of genetic diversity depend on genealogy, we can develop inference procedures to learn about the past.

The most commonly used genealogical model is the Kingman coalescent, which arises in many models of neutral evolution~\autocite{Kingman1982, Hudson1983, Tajima1983}.
The key characteristic of the Kingman coalescent is that only two lineages are permitted to share a common ancestor at the same point in time.
Moreover, the genealogies generated by the Kingman coalescent are randomly bifurcating trees: each node subtends exactly two branches and the number of leaves subtended by one of these branches is uniformly distributed~\autocite{Tajima1983}.

% Kingman coalescent models are appropriate when a single timescale controls patterns of relatedness between individuals in a sample from a population.
% This timescale, which we will call $T_2$, is inversely proportional to the rate of coalescence between pairs of lineages and sets the branch lengths of genealogies.
% Thus, $T_2$ determines observable quantities such as the average genetic diversity between pairs of individuals and length distribution of genomic segments that are identical-by-descent.
% Population genetic inference methods interpret these quantities in light of the Kingman distribution of tree topologies to estimate the coalescent timescale.
% In the simplest models of neutral evolution, the coalescent timescale is proportional to the number of individuals in the population. Therefore, it is commonly referred to as the effective population size.

%The Kingman coalescent generates genealogies with several distinguishing characteristics. First, the genealogies are binary trees: each node has only two children. A second and related topological feature is that for a node subtending $n$ leaves, the number of leaves subtended by one of its children is uniformly distributed between 1 and $n-1$. Finally, under a neutral model of evolution, the rate of coalescence (and thus the branch lengths of the genealogy) is inversely proportional to the population size.

Because computing the full likelihood of the data is generally intractable, population genetic inference is typically done on informative summary statistics.
One informative statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in a sample.
The expected number of mutations in $i$ sampled chromosomes is proportional to the lengths of branches subtending $i$ leaves of the genealogy, and thus depends the distributions of topologies as well as branch lengths.
By assuming the Kingman model, one can marginalize over the unobserved tree topologies and extract information about the distribution of branch lengths and thus about the rate of coalescence as a function of time.
If this assumption is good, the site frequency spectrum thus reflects the history of population size changes.
For example, high-frequency alleles are the result of mutations on deep branches and their number in a sample reflects the population size in the distant past.

A serious problem for this inference procedure is that alternative models of evolution generate different coalescent models.
For example, models of pervasive weak selection, skewed offspring number distributions, and periodic strong bottlenecks can all generate genealogies that differ from the Kingman both topologically and in terms of branch length.
[Review of MMC literature here? Neutral models (Eldon, Birkner, etc.), sweeps (Durett Schweinsberg, Coop and Ralph), rapid adaptation (Neher, Desai), background selection (Seger, Good, Nicholaisen), Range expansion (??)]
Collectively, these models are known as \emph{multiple mergers coalescents} because, unlike the Kingman coalescent, they permit multiple lineages to coalesce instantaneously.
As a result, multiple-mergers genealogies contain nodes with more than two children and lack the uniform branching structure of the Kingman.
Even more important for inference is the fact that the coalescent timescales---and thus the branch lengths and levels of diversity---are determined by parameters other than the population size.
In this case, the term `effective population size' is misleading because $T_2$ is no longer proportional to the number of individuals in the population.
Thus, the results of a Kingman-based inference procedure applied to data generated by a non-Kingman process will be both quantitatively and qualitatively wrong.
Clearly, it would be useful to be able to distinguish among different coalescent models, both to check that popular inferences of historical population sizes are valid and to understand which evolutionary forces are dominant in various populations.

[Mention the distinction between single-locus models (e.g. Seger on Whale lice, Neher et al. with viruses) where you can explicitly infer trees, and multi-locus methods, where you average over many latent genealogies. This method is in the latter group.]

An obvious question is whether we can use the average site frequency spectrum to reject multiple-mergers.
There are two distinguishing features of the SFS generated by multiple mergers coalescents.
First, low-frequency mutations are enriched relative to the time-homogeneous Kingman coalescent.
Unfortunately, this is also a feature of the Kingman coalescent in growing populations, so it is not well-suited for model selection (but see \cite{}).
Second, the multiple-merger SFS is non-monotonic: it has a positive slope for high frequencies.
Unlike the excess of rare alleles, this feature cannot be reproduced by an exchangeable Kingman coalescent model.
However, it is difficult to identify the non-monotonicity in real data.
The problem is that it necessary to know which allele is ancestral to distinguish derived mutations at frequency $x$ from those at frequency $1-x$.
Various methods for identifying the ancestral allele exist, but in practice even a very low rate of error can generate a spurious signal.

Ideally, we would like a statistic (or statistics) that:
\begin{enumerate}
\item is qualitatively different between models of genealogies,
\item is robust to demographic changes, and
\item requires minimal information beyond genetic diversity.
\end{enumerate}
The first two criteria are important for making qualitative distinctions between models. The last one would maximize its applicability.
Here, I propose that the joint site frequency spectrum at pairs of linked sites is such a statistic.

\section*{Single-locus theory}

\section*{Recombination}

@@

>>>>

One of the most important and widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
The Kingman coalescent is a stochastic process that generates gene genealogies: trees representing the patterns of shared ancestry of sampled individuals.
Inference methods use these genealogies as latent variables linking demographic parameters to observable data~\autocite{RosenbergNordborg2002}.
The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002})---tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).

In its simplest form, the Kingman coalescent has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
Under a variety of conditions, the coalescent rate is inversely proportional to the size of the population~\autocite{Kingman1982b}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
This observation is the basis for a variety of methods to infer historical population sizes from genetic data.

With modern population genomics data sets, full-data likelihood models are impractical, so population-size inference is typically done on informative summary statistics.
One widely used statistic is the site frequency spectrum (SFS): the number of mutations observed as a function of their allele frequency in the sample.
The expected number of mutations at a given frequency depends on the branch lengths, and hence, on the coalescent rate.
Thus, one can infer the coalescent rate by integrating over tree topologies, weighted by their probabilities under the Kingman coalescent.
Some methods perform this integration by Monte Carlo simulation (e.g., \cite{CoventryEtAl2010, ExcoffierEtAl2013}).
Others (e.g., \cite{Nielsen2000}, \cite{BhaskarEtAl2015}), compute the expected site frequency spectrum directly for simple demographic models using the results of \cite{GriffithsTavare1998} or \cite{PolanskiKimmel2003}.
(Another class of SFS-based methods are based on corresponding forward-time models rather than the coalescent \autocite{GutenkunstEtAl2009, LukicEtAl2011, RagsdaleGutenkunst2017, JouganousEtAl2017}.)

A serious problem for this inference procedure is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
When this assumption is violated, Kingman-based inference methods are misspecified.
For example, \cite{SchriderEtAl2016} recently demonstrated how population size inference can be distorted by selective sweeps.
This effect is present in SFS-based methods as well as in sequentially Markov coalescent methods (e.g., \cite{LiDurbin2011}).
In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce overall genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
Genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of neutral and non-neutral contexts.
Whereas in the Kingman coalescent lineages may only coalesce pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph},
rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.
In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps \autocite{DurrettSchweinsberg2005}.
In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be \emph{qualitatively} incorrect.
It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.

One possible approach to identifying multiple mergers in genomic data to use the SFS as a summary statistic.
(\cite{Koskela2015} developed a full-likelihood method based on importance sampling, but as with other ``exact'' inference methods it does not scale to genomic data.)
\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merge coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected non-Kingman signatures of widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatscheck2013, BirknerEtAl2013}.

However, existing methods have limitations for distinguishing multiple mergers from general models of population-size change.
The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
\cite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers with sufficient data, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but detecting it in data requires knowledge of the ancestral allele at each site.
High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS at high frequencies.

Here, we propose that summary statistics based on the spectrum of mutation frequencies at pairs of nearby site---the 2-SFS \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
Finally, we demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

\section*{Definitions and single-locus theory}
[Define statistic, show that it works without recombination. (Detects multiple mergers, doesn't respond to population size changes)]

\section*{Simulations with recombination}
[Coalescent simulations with recombination (it still works).]

\section*{Linked selective sweeps}
[Forward-time simulations of linked sweeps (still works).]

@@

\section*{Discussion}
\begin{itemize}
    \item Focus here on coalescent, but also relevant for diffusion methods
    \item Single-locus tests that rely on trees (Seger whale lice, Neher with viruses)
    \item Multi-locus tests that compare vs genomic background
    \item Could extend to a local method to look for variation in ``multiple-merger-ness''
    \item Could also extend
\end{itemize}

Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).
